---
title: "MPHBioRevision"
author: "YunhuiQi"
date: "2026-01-13"
output: html_document
---

```{r packages}
library(edgeR)              
library(limma)              
library(ggplot2)            
library(ggfortify)          
library(variancePartition)  
library(ggpubr)      
library(stringr)
library(dplyr)
library(tidyverse)
library(tidyverse)
library(pROC)
library(DESeq2)
library(MASS)
library(nleqslv)
library(rootSolve)
library(pracma)
library(ComplexHeatmap)
library(pheatmap)
library(plotly)
library(cowplot)
library(DescTools)
library(glmmTMB)
library(emmeans)
library(parallel)
```


```{r functions for 2sLRT}
Hetero_NB_LRT <- function(y_g1, y_g2, y_g3, C_1 = 1, C_2 = 2, C_3 = 1, dispersion = 1){
  # This function conducts NB LRT test for heterosis in a single family and for a single gene with single replication per genotype given dispersion parameter
  # Hg0: mu_g1 + mu_g2 = 2 * mu_g3
  # @param y_g1: raw count for female
  # @param y_g2: raw count for male
  # @param y_g3: raw count for hybrid
  # @param C_1: normalization factor for female sample
  # @param C_2: normalization factor for male sample
  # @param C_3: normalization factor for hybrid sample
  # @param dispersion: estimation of dispersion parameter
  # @output pval: pvalue for NB LRT test with known dispersion
  
  # rate = 1/dispersion
  if(dispersion == 0){return(NA)}
  r <- 1/dispersion
  
  # 1. Get MLE under global
  mu_g_hat1 <- c(y_g1/C_1, y_g2/C_2, y_g3/C_3)
  
  # 2. Get MLE under null
  y_set <- c(y_g1, y_g2, y_g3)
  C_set <- c(C_1, C_2, C_3)
  if(sum(y_set) == 0){return(1)}
  else{
    equations <- function(mu_set) {
      eq1 <- y_set[1]/mu_set[1] + y_set[3]/(mu_set[1]+mu_set[2]) - C_1*(r + y_set[1])/(C_1*mu_set[1] + r) - C_3*(r + y_set[3])/(C_3*(mu_set[1] + mu_set[2]) + 2*r)
      eq2 <- y_set[2]/mu_set[2] + y_set[3]/(mu_set[1]+mu_set[2]) - C_2*(r + y_set[2])/(C_2*mu_set[2] + r) - C_3*(r + y_set[3])/(C_3*(mu_set[1] + mu_set[2]) + 2*r)
      return(c(eq1, eq2))
    }
    
    init_list <- list(c(ifelse(y_g1 == 0, 10, y_g1), ifelse(y_g2 == 0, 10, y_g2)),
                      c(y_g1/C_1, y_g2/C_2), c(y_g1/2, y_g2/2),
                      c(10, 10), c(100, 100), c(2, 2), c(100, 10), c(10, 100),
                      c(100, 2), c(2, 100))

    # function to check the validity of the solution
    check_solution <- function(sol) {
      all(sol$x >= 0) && all(sol$x <= max(c(y_g1, y_g2, y_g3)) * 100) && sol$termcd < 4
    }
    
    # loop through the starting points to find a valid solution
    sol_found <- FALSE
    for (inits in init_list) {
      solution <- nleqslv(x = inits, fn = equations, control = list(maxit = 500))
      if (check_solution(solution)) {
        mu_g_hat0 <- c(solution$x, mean(solution$x))
        sol_found <- TRUE
        break
      }
    }
    
    # if no solution is found with nleqslv, use multiroot
    if (!sol_found) {
      root <- multiroot(f = equations, start = c(ifelse(y_g1 == 0, 10, y_g1/C_1), ifelse(y_g2 == 0, 10, y_g2/C_2)))
      if (all(root$root >= 0) && all(root$root <= max(c(y_g1, y_g2, y_g3)) * 100)) {
        mu_g_hat0 <- c(root$root, mean(root$root))
      } 
      else {
        return(NA) # Or handle the failure to find a solution as appropriate
      }
    }

    logL <- function(y_set, C_set, mu_set, r){
      # under H1, if y_gi!=0, mu_gi!=0, under H0, its very rare to get 0 as MLE
      if(mu_set[1] == 0){first <- -(r + y_set[1])*log(r + C_set[1]*mu_set[1])}
      else{first <- y_set[1]*log(mu_set[1]) - (r + y_set[1])*log(r + C_set[1]*mu_set[1])}
      if(mu_set[2] == 0){second <- -(r + y_set[2])*log(r + C_set[2]*mu_set[2])}
      else{second <- y_set[2]*log(mu_set[2]) - (r + y_set[2])*log(r + C_set[2]*mu_set[2])}
      if(mu_set[3] == 0){third <- -(r + y_set[3])*log(r + C_set[3]*mu_set[3])}
      else{third <- y_set[3]*log(mu_set[3]) - (r + y_set[3])*log(r + C_set[3]*mu_set[3])}
      return(first + second + third)
      # Since the choose terms and rlogr and ylogC terms will cancel even with y=0, but it can be greater than the computation can afford  to make computation fail, so we omit them right now
  }
  LR <- -2 * (logL(y_set=y_set, C_set = C_set, mu_set = mu_g_hat0, r=r) - logL(y_set=y_set, C_set = C_set, mu_set = mu_g_hat1, r=r))
  pval <- pchisq(LR, df = 1, lower.tail = FALSE)
  # return(list(pval=pval, mu_g_hat0 = mu_g1_hat0, mu_g_hat1 = mu_g_hat1))
  return(pval)
  }
}
Pois_3mean_LRT <- function(y_g1, y_g2, y_g3, C_1 = 1, C_2 = 1, C_3 = 1){
  # This function conducts Poisson LRT test for non-family in a single family and for a single gene with single replication per genotype
  # Hg0: mu_g1 = mu_g2 = mu_g3
  # @param y_g1: raw count for female
  # @param y_g2: raw count for male
  # @param y_g3: raw count for hybrid
  # @param C_1: normalization factor for female sample
  # @param C_2: normalization factor for male sample
  # @param C_3: normalization factor for hybrid sample
  
  y_set <- c(y_g1, y_g2, y_g3)
  C_set <- c(C_1, C_2, C_3)
  mu_g_hat0 <- c(sum(y_set)/sum(C_set), sum(y_set)/sum(C_set), sum(y_set)/sum(C_set))
  mu_g_hat1 <- c(y_g1/C_1, y_g2/C_2, y_g3/C_3)
  logL <- function(y_set, C_set, mu_set){
    # if y_gi != 0, mu_gi != 0, for i=1,2,3
    if(mu_set[1] == 0){first <- 0}
    else{first <- y_set[1]*log(mu_set[1]) - C_set[1]*mu_set[1]}
    if(mu_set[2] == 0){second <- 0 }
    else{second <- y_set[2]*log(mu_set[2]) - C_set[2]*mu_set[2]}
    if(mu_set[3] == 0){third <- 0}
    else{third <- y_set[3]*log(mu_set[3]) - C_set[3]*mu_set[3]}
    return(first + second + third)
    #- log(factorial(y_set[1])) - log(factorial(y_set[2])) - log(factorial(y_set[3]))
    #+ y_set[1]log(C_set[1]) + y_set[2]log(C_set[2]) + y_set[3]log(C_set[3]) 
    # Since the factorial and ylogC terms will cancel even when y=0, but it can be infinity to make computation fail, so we omit them.
  }
  LR <- -2 * (logL(y_set = y_set, C_set = C_set, mu_set = mu_g_hat0) - logL(y_set = y_set, C_set = C_set, mu_set = mu_g_hat1))
  pval <- pchisq(LR, df = 2, lower.tail = FALSE)
  # return(list(pval=pval, mu_g_hat0 = mu_g1_hat0, mu_g_hat1 = mu_g_hat1))
  return(pval)
}
Disp_est_null_family <- function(counts, group = NULL){
  # This function estimate the tagwise dispersion parameter for multiple genes with multiple family
  # @param counts: with genes in rows, and three*n_families columns, female, male and hybrid which passed non-family LRT
  # @param group: a vector indicating the group information of three*n_families columns
  if(is.null(group)){
    y <- DGEList(counts = counts)
    y <- calcNormFactors(y, method = "RLE")
    y <- estimateCommonDisp(y)
    y <- estimateTrendedDisp(y)
    y <- estimateTagwiseDisp(y)
  }
  else{
    y <- DGEList(counts = counts, group = group) 
    y <- calcNormFactors(y, method = "RLE")
    design <- model.matrix(~ group)
    y <- estimateGLMCommonDisp(y, design)
    y <- estimateGLMTrendedDisp(y, design)
    y <- estimateGLMTagwiseDisp(y, design)
  }
    return(y$tagwise.dispersion)
  }

```

# Comparison using Pitz replicated dataset

## 1. Data cleaning

```{r}
meta_all <- read.table("sample_info_v2.txt", header = TRUE, sep = "\t", fill=TRUE, row.names = "SampleName")
counts_gen <- read.table("filtered_counts_834_samples.txt",, header = TRUE, sep = "\t", fill=TRUE, row.names = 1)
all(sort(colnames(counts_gen))==sort(rownames(meta_all)[meta_all$keep_exclude=="keep"]))

counts_pitz <- read.csv("PE_Reads_Table_Gen_pitz.csv",sep = ";", row.names = "Gene_ID")
colnames(counts_pitz) <- rownames(meta_all)[match(colnames(counts_pitz), meta_all$Sample)]

counts <- counts_pitz[,colnames(counts_gen)]
rm(counts_pitz, counts_gen)

metadata <- meta_all[colnames(counts),]
metadata$Batch <- as.factor(metadata$Batch)
metadata$System <- as.factor(metadata$System)
metadata$Replicate <- as.factor(metadata$Replicate.Block)
metadata$Genotype <- as.factor(gsub(" ", "", metadata$Genotype))
metadata$REP.BAT <- factor(paste(metadata$Replicate, metadata$Batch, sep = "_"))
metadata$REP.BAT.SYS <- factor(paste(metadata$Replicate, metadata$Batch, metadata$System, sep = "_"))


DEGs_pitz <- list.files("/Users/yunhuiqi/Library/Mobile Documents/com~apple~CloudDocs/PhD/academic/2023 Spring/RA/MazeHeterosis/BioPaperRevision/DEGs_LFC1_p05/", pattern = "\\.csv$", full.names = TRUE)
DEGs_pitz <- setNames(lapply(DEGs_pitz, read.csv, row.names = 1),  sub(".*_([^_]+_[^_]+)\\.csv$", "\\1", basename(DEGs_pitz)))
DEGs_pitz <- lapply(DEGs_pitz, function(x){x[x$Additivity == "non-additive",]})
```


## 2. Batch and system variation assessment

Evaluation of Technical Noise and Variance Partitioning
Raw gene count matrices were pre-processed using the edgeR package (Robinson et al., 2010). To ensure consistency with the established analytical framework of the study, we applied the normalization and filtering criteria described by Pitz et al. (2025). Specifically, normalization factors were calculated using the Trimmed Mean of M-values (TMM) method. Lowly expressed genes were subsequently filtered using the filterByExpr function; crucial to this step, we utilized the specific design matrix defined by Pitz et al. (~ 0 + Genotype + REP.BAT), which accounts for the incomplete block design by incorporating the batch factor nested within replicates.

To quantify the contribution of technical covariates to the total gene expression variability, we performed a Variance Component Analysis (VCA) using the variancePartition package (Hoffman and Schadt, 2016). While Pitz et al. modeled the nested "System" effect using duplicateCorrelation within a linear model for differential expression (Eqn 1 in Pitz et al., 2025), we adapted their hierarchical model definition to explicitly partition variance components. The mixed-effects model was specified as expression ~ (1|Replicate) + (1|REP.BAT) + (1|REP.BAT.SYS). This formula mirrors the alpha-design structure described in Pitz et al., where growth systems are nested within batches (REP.BAT.SYS) and batches are nested within replicates (REP.BAT). By treating these hierarchical layers as random effects, we were able to calculate the fraction of variance explained by biological replication versus technical processing layers (Batch and System) for every gene.

```{r}
#' Evaluate Technical Noise (Pitz et al. Method)
#'
#' @param counts_all Raw integer count matrix (Genes x All Samples).
#' @param meta_all Metadata dataframe (Rows match counts_all columns).
#' @param target_idx Integer vector indicating which samples to analyze (e.g., Rep 1).
#' @param vp_formula Formula for variance partition. IMPORTANT: Use the nested names! 
#'                   e.g., ~ (1|REP.BAT) + (1|REP.BAT.SYS)
#' @param plot_title String. Label for plots.
#'
#' @return List containing plots, stats, and normalized data.
evaluate_technical_noise <- function(counts_all, 
                                     meta_all, 
                                     target_idx, 
                                     vp_formula, 
                                     plot_title = "Analysis") {
  

  # ============================================================================
  # Global Normalization 
  # ============================================================================
  # Justification: Matches Author's Step 1. 
  # Calculate scaling factors on the full dataset before removing genes.
  message(">>> Phase 1: Global Normalization (TMM)...")
  
  dge <- DGEList(counts = counts_all)
  dge <- calcNormFactors(dge, method = "TMM")

  
  # ============================================================================
  # Filtering using Design Matrix
  # ============================================================================
  # Justification: Matches Author's Step 2 & 3.
  # Using the design matrix (~ Genotype + REP.BAT) is the most robust way 
  # to filter genes in complex designs.
  message(">>> Phase 2: Global Filtering using Pitz Design...")
  
  # Construct design matrix dynamically
  # Pitz used: ~ 0 + Genotype + REP.BAT
  design_str <- "~ 0"
  if("Genotype" %in% colnames(meta_all)) design_str <- paste(design_str, "+ Genotype")
  # Important: Use the NESTED batch variable we created
  design_str <- paste(design_str, "+ REP.BAT")
  
  message("    Design formula: ", design_str)
  design <- model.matrix(as.formula(design_str), data = meta_all)
  
  # Filter
  keep <- filterByExpr(dge, design)
  dge_clean <- dge[keep, , keep.lib.sizes = FALSE]
  
  message("    Original Genes: ", nrow(counts_all))
  message("    Retained Genes: ", sum(keep))
  
  # ============================================================================
  # PHASE 3: Generate Data & Subset
  # ============================================================================
  message(">>> Phase 3: Generating Data & Subsetting...")
  
  # Calculate CPM on the CLEAN data
  # Log2-CPM for Stats/Plots
  log_data_all <- cpm(dge_clean, normalized.lib.sizes = TRUE, log = TRUE, prior.count = 0.5)
  # Linear CPM for Output
  linear_data_all <- cpm(dge_clean, normalized.lib.sizes = TRUE, log = FALSE)
  
  # Now Subset to the Target (e.g., B73 or Replicate 1)
  data_sub_log <- log_data_all[, target_idx]
  data_sub_linear <- linear_data_all[, target_idx]
  meta_sub <- meta_all[target_idx, ]
  
  # Safety: Drop unused factor levels in the subset
  meta_sub <- droplevels(meta_sub)
  
  # Safety: Remove genes with 0 variance inside this specific subset (prevents PCA crash)
  gene_vars <- apply(data_sub_log, 1, var)
  keep_safe <- gene_vars > 1e-6
  data_clean_log <- data_sub_log[keep_safe, ]
  data_clean_linear <- data_sub_linear[keep_safe, ]
  
  # ============================================================================
  # PHASE 4: PCA
  # ============================================================================
  message(">>> Phase 4: PCA...")
  
  pca_res <- prcomp(t(data_clean_log), center = TRUE, scale. = FALSE)
  pct_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)
  
  # Plot using the NESTED variables if possible, or standard Batch/System labels
  # Note: In Rep 1, "REP.BAT" and "Batch" usually look identical visually
  p_pca_batch <- autoplot(pca_res, data = meta_sub, colour = "Batch", size = 3) +
    labs(title = paste0("PCA: ", plot_title),
         subtitle = "Color: Batch",
         x = paste0("PC1: ", pct_var[1], "%"), y = paste0("PC2: ", pct_var[2], "%")) +
    theme_bw() + theme(legend.position = "bottom")
  
  p_pca_system <- autoplot(pca_res, data = meta_sub, colour = "System", size = 3) +
    labs(title = paste0("PCA: ", plot_title),
         subtitle = "Color: System",
         x = paste0("PC1: ", pct_var[1], "%"), y = paste0("PC2: ", pct_var[2], "%")) +
    theme_bw() + theme(legend.position = "bottom")
  
  p_pca <- ggarrange(p_pca_batch, p_pca_system, nrow = 1)
  
  # ============================================================================
  # PHASE 5: Variance Partition
  # ============================================================================
  message(">>> Phase 5: Variance Partition...")
  
  # Fit Model
  vp_res <- fitExtractVarPartModel(data_clean_log, vp_formula, meta_sub)
  
  # Calculate summary stats to help decision making
  vp_stats <- apply(vp_res, 2, median) * 100
  
  p_vp <- plotVarPart(sortCols(vp_res)) +
    labs(title = paste0("Variance Partition: ", plot_title),
         subtitle = paste0("Median Batch Effect (REP.BAT): ", 
                           round(vp_stats[grep("BAT", names(vp_stats))[1]], 2), "%")) +
    theme_bw()
  
  message(">>> Done.")
  
  return(list(
    pca_plot = p_pca,
    vp_plot = p_vp,
    vp_results = vp_res,
    vp_stats = vp_stats,
    norm_counts = data_clean_linear,
    log_trans_data = data_clean_log,
    meta_subset = meta_sub
  ))
}
```
```{r}
# 1. All replicates, B73
eval_B73_all_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "B73"), 
                         vp_formula = ~ (1|Replicate) + (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "B73 - All replicates - TMM")

# 2. All replicates, Mo17
eval_Mo17_all_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "Mo17"), 
                         vp_formula = ~ (1|Replicate) + (1|REP.BAT) + (1|REP.BAT.SYS),
                         plot_title = "Mo17 - All replicates - TMM")

# 3. Rep 1, B73
eval_B73_Rep1_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "B73" & metadata$Replicate == "1"), 
                         vp_formula = ~ (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "B73 - Rep 1 - TMM")

# 4.  Rep 1, Mo17
eval_Mo17_Rep1_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "Mo17" & metadata$Replicate == "1"), 
                         vp_formula =  ~ (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "Mo17 - Rep 1 - TMM")


# 5. Rep 2, B73
eval_B73_Rep2_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "B73" & metadata$Replicate == "2"), 
                         vp_formula =  ~ (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "B73 - Rep 2 - TMM")

# 6.  Rep 2, Mo17
eval_Mo17_Rep2_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "Mo17" & metadata$Replicate == "2"), 
                         vp_formula =  ~ (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "Mo17 - Rep 2 - TMM")

# 7. Rep 3, B73
eval_B73_Rep3_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "B73" & metadata$Replicate == "3"), 
                         vp_formula =  ~ (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "B73 - Rep 3 - TMM")

# 8.  Rep 3, Mo17
eval_Mo17_Rep3_TMM <- evaluate_technical_noise(counts_all = counts, 
                         meta_all = metadata,
                         target_idx = which(metadata$Genotype == "Mo17" & metadata$Replicate == "3"), 
                         vp_formula =  ~ (1|REP.BAT) + (1|REP.BAT.SYS), 
                         plot_title = "Mo17 - Rep 3 - TMM")

```
```{r}
round(eval_B73_all_TMM$vp_stats,4)
round(eval_Mo17_all_TMM$vp_stats,4)
```


## 3. Full family search

To validate our single-replicate method without confounding technical variation, we restricted our analysis to "spatially matched trios" where the hybrid and both parents were processed in the same experimental Batch and grown in the exact same aeroponic System. This strict filtering eliminates temporal and spatial noise, ensuring MPH calculations reflect biological differences. We identified a total of 169 such families available for high-precision validation (39 in Replicate 1, 65 in Replicate 2, and 65 in Replicate 3)

```{r}
# 1. Initialize an empty list to store the results
# Using a list is faster than repeatedly using rbind inside a loop
match_list <- list()
counter <- 1

# 2. Ensure Genotype column is character type (not factor) to avoid errors
# metadata$Genotype <- as.character(metadata$Genotype)

metadata$SampleName <- rownames(metadata)
sample_id_col <- "SampleName" 

# 3. Loop through each unique Biological Replicate
for (rep in unique(metadata$Replicate)) {
  
  # Subset data for the current replicate
  rep_data <- metadata[metadata$Replicate == rep, ]
  
  # Identify Hybrid genotypes: contain an underscore "_"
  all_gts <- unique(rep_data$Genotype)
  hyb_gts <- grep("_", all_gts, value = TRUE)

  # Loop through each Hybrid Genotype found in this replicate
  for (hyb in hyb_gts) {
    
    # Extract Parent names from the string (assuming format "Mother_Father")
    parts <- strsplit(hyb, "_")[[1]]
    mom_name <- trimws(parts[1]) # Parent 1
    dad_name <- trimws(parts[2]) # Parent 2
    
    # Get all rows for this specific Hybrid genotype in this replicate
    hyb_rows <- rep_data[rep_data$Genotype == hyb, ]
    
    # Loop through each individual Hybrid sample (row)
    for (i in 1:nrow(hyb_rows)) {
      
      # Get specific info for the current hybrid sample
      curr_hyb_sample <- hyb_rows[i, sample_id_col]
      curr_batch <- hyb_rows$Batch[i]
      curr_system <- hyb_rows$System[i]
      
      # ---------------------------------------------------------
      # CORE LOGIC: Find parents in the SAME Replicate, Batch, and System
      # ---------------------------------------------------------
      
      # Find Mother (Parent 1) matches
      mom_matches <- rep_data[rep_data$Genotype == mom_name & 
                              rep_data$Batch == curr_batch & 
                              rep_data$System == curr_system, ]
      
      # Find Father (Parent 2) matches
      dad_matches <- rep_data[rep_data$Genotype == dad_name & 
                              rep_data$Batch == curr_batch & 
                              rep_data$System == curr_system, ]
      
      # ---------------------------------------------------------
      # Store Data
      # ---------------------------------------------------------
      
      match_list[[counter]] <- data.frame(
        Replicate = rep,
        
        # Hybrid Info
        Hybrid_Genotype = hyb,
        Hybrid_Sample = curr_hyb_sample,
        Batch = curr_batch,
        System = curr_system,
        
        # Parent 1 (Mother) Info
        Parent1_Genotype = mom_name,
        Parent1_Count = nrow(mom_matches), # How many matching samples found
        # Paste sample names together if there are multiple matches
        Parent1_Samples = paste(mom_matches[, sample_id_col], collapse = "; "), 
        
        # Parent 2 (Father) Info
        Parent2_Genotype = dad_name,
        Parent2_Count = nrow(dad_matches),
        Parent2_Samples = paste(dad_matches[, sample_id_col], collapse = "; "),
        
        # Logic Check: TRUE if both parents are found in the same batch/system
        Full_Match_Found = (nrow(mom_matches) > 0 & nrow(dad_matches) > 0)
      )
      
      counter <- counter + 1
    }
  }
}

# 4. Combine the list into a single Data Frame
final_tracking_table <- do.call(rbind, match_list)

# 5. Preview results
meta_family <- final_tracking_table[final_tracking_table$Full_Match_Found == TRUE, ] %>% 
  mutate(
    Parent1_Samples_Single = str_remove(Parent1_Samples, ";.*"),
    Parent2_Samples_Single = str_remove(Parent2_Samples, ";.*")
  )
table(meta_family$Replicate)
```





## 4. Apply two stage LRT to a single replicate 


For each independent replicate run, we first subsetted the raw count matrix to include only the samples corresponding to the identified spatially matched families (Hybrid, Maternal Parent, Paternal Parent). A pseudo-count of +1 was added to all raw counts to ensure numerical stability in downstream algorithms. We then performed standard preprocessing using the edgeR package: raw counts were normalized using the Trimmed Mean of M-values (TMM) method to account for sequencing depth, and lowly expressed genes were filtered using the filterByExpr function based on Genotype grouping

```{r}
# ------------------------------------------------------------------------------
# STEP 1: SUBSET RAW DATA (Crucial Step)
# ------------------------------------------------------------------------------
# Isolate single replicate samples from your raw counts and metadata
# 'target_idx' contains sample names for hybrids in this replicate with its parents are in the same batch and system
rep_selected <- "1"
meta_family_rep <- meta_family[meta_family$Replicate == rep_selected, ]
hybrid_info <- data.frame(hybrid = meta_family_rep$Hybrid_Sample,
                          female = meta_family_rep$Parent1_Samples_Single,
                          male = meta_family_rep$Parent2_Samples_Single)
counts_rep <- counts[, unique(c(hybrid_info$hybrid, hybrid_info$female, hybrid_info$male))] + 1 # we add pseudo count 1 to avoid 0 which will make our algorithm fail.
meta_rep   <- metadata[colnames(counts_rep), ]

# ------------------------------------------------------------------------------
# STEP 2: INITIALIZE & NORMALIZE (Pitz Step 1)
# ------------------------------------------------------------------------------
# Create DGEList for the single replicate
# Use Genotype as the group
dge_rep <- DGEList(counts = counts_rep, group = meta_rep$Genotype)

# Run TMM Normalization, just like Pitz et al.
dge_rep <- calcNormFactors(dge_rep, method = "TMM")

# ------------------------------------------------------------------------------
# STEP 3: FILTERING (Pitz Step 2)
# ------------------------------------------------------------------------------
# We simply use the group (Genotype) logic which filterByExpr handles automatically.

rep_keep <- filterByExpr(dge_rep, group = meta_rep$Genotype)

# Apply the filter
# keep.lib.sizes=FALSE ensures library sizes are refreshed after removing genes
dge_rep_clean <- dge_rep[rep_keep, , keep.lib.sizes = FALSE]

message("Original Genes: ", nrow(counts_rep))
message("Retained Genes: ", nrow(dge_rep_clean))

# ------------------------------------------------------------------------------
# STEP 4: GENERATE NORMALIZED DATA
# ------------------------------------------------------------------------------
# Get TMM-Normalized CPM (Counts Per Million)
# This is the input for your method
norm_data_rep <- cpm(dge_rep_clean, normalized.lib.sizes = TRUE, log = FALSE)




raw_use <- dge_rep_clean$counts # this is the pseudo count
norm_use <- norm_data_rep
norm_factors_use <- dge_rep_clean$samples$norm.factors
names(norm_factors_use) <- rownames(dge_rep_clean$samples)
```


```{r perform 2sLRT on real data, eval = FALSE}
##### Step 1: Gene-wise pre-test - Poisson LRT for null-family ####
  gene_test_func <- function(gene){
  # compute pvalues from test using raw counts and norm factors
    pval_family <- c()  # for Poisson LRT equal family
    if(sd(gene) == 0){
      # if this gene is equally expressed in all varieties
      pval_family <- rep(1, nrow(hybrid_info))
    }
    for(hb_num in 1:nrow(hybrid_info)){
      female <- as.numeric(gene[hybrid_info$female[hb_num]])
      male <- as.numeric(gene[hybrid_info$male[hb_num]])
      hybrid <- as.numeric(gene[hybrid_info$hybrid[hb_num]])
      
      pval_family[hb_num] <- Pois_3mean_LRT(y_g1 = female, y_g2 = male, y_g3 = hybrid,
                                            C_1 = norm_factors_use[hybrid_info$female[hb_num]],
                                            C_2 = norm_factors_use[hybrid_info$male[hb_num]],
                                            C_3 = norm_factors_use[hybrid_info$hybrid[hb_num]])
      names(pval_family)[hb_num] <- hybrid_info$hybrid[hb_num]
    }
    return(list(pval_family = pval_family))
}

  gene_est_func <- function(gene){
  # get MP, MPH ratio and diff MPH ratio from normalized data, i.e. estimation of mu_gi under global/Hg1
    MP <- c()           # mid parents
    hyb_ratio <- c()    # MPH ratio = hyb/MP
    diff_ratio <- c()   # MPH diff ratio = (hyb-MP) / MP
    family_mean <- c()  # family mean 
    if(all(gene == 0)){
      # if this gene is not expressed in all varieties
      MP <- hyb_ratio <- diff_ratio <- family_mean <- rep(0, nrow(hybrid_info))
    }
    for(hb_num in 1:nrow(hybrid_info)){
      female <- as.numeric(gene[hybrid_info$female[hb_num]])
      male <- as.numeric(gene[hybrid_info$male[hb_num]])
      hybrid <- as.numeric(gene[hybrid_info$hybrid[hb_num]])
      
      MP[hb_num] <- mean(c(female, male))
      family_mean[hb_num] <- mean(c(female, male, hybrid))
      hyb_ratio[hb_num] <- hybrid/(MP[hb_num])
      diff_ratio[hb_num] <- (hybrid - MP[hb_num]) / (MP[hb_num])
      names(MP)[hb_num] <- names(family_mean) <- names(hyb_ratio)[hb_num] <- names(diff_ratio)[hb_num] <- hybrid_info$hybrid[hb_num]
    }
    return(list(MP = MP, family_mean = family_mean, hyb_ratio = hyb_ratio, diff_ratio = diff_ratio))
  }
  
  test_result_list <- lapply(1:nrow(raw_use), function(i) gene_test_func(raw_use[i, ]))
  est_result_list <- lapply(1:nrow(norm_use), function(i) gene_est_func(norm_use[i, ]))

  
  # combine genewise pre-test as a matrix, the order of results is consistent of hybrids in hybrid_info.
  pval_family <- do.call(rbind, lapply(test_result_list, function(x) x$pval_family))
  MP <- do.call(rbind, lapply(est_result_list, function(x) x$MP))
  family_mean <- do.call(rbind, lapply(est_result_list, function(x) x$family_mean))
  diff_ratio <- do.call(rbind, lapply(est_result_list, function(x) x$diff_ratio))
  hyb_ratio <- do.call(rbind, lapply(est_result_list, function(x) x$hyb_ratio))
  rownames(pval_family) <- rownames(MP) <- rownames(hyb_ratio) <- rownames(diff_ratio) <- rownames(family_mean) <- rownames(raw_use)
  colnames(pval_family) <- colnames(MP) <- colnames(hyb_ratio) <- colnames(diff_ratio) <- colnames(family_mean) <- hybrid_info$hybrid

  qval_family <- apply(pval_family, 2, function(x){p.adjust(x, method = "fdr")})
  

##### Step 2: Estimate dispersion matrix ####
  # values of tuning parameters
  gene_groups_no = 1     # number of clusters
  n_fam_thres = 15       # number of null families threshold for dispersion estimation
  alpha0 <- 0.1

  null_disp_ind <- qval_family > alpha0
  DISP <- rep(NA, nrow(null_disp_ind)) # each gene will only have one dispersion estimation
  names(DISP) <- rownames(null_disp_ind)
  null_disp_ind_sub <- null_disp_ind[apply(null_disp_ind, 1, function(x) sum(x == 1)) >= n_fam_thres,]
  jacc_dist <- dist(null_disp_ind_sub, "binary")
  jacc_hc <- hclust(jacc_dist)
  jacc_clust <- cutree(jacc_hc, gene_groups_no)
  cat("Gene clusters size: \n")
  print(table(jacc_clust))
  for (i_clust in unique(jacc_clust)) {
    genes_clust <- rownames(null_disp_ind_sub)[jacc_clust == i_clust]
    ind_clust <- null_disp_ind_sub[genes_clust,]
    worst_gene <- genes_clust[which.max(apply(ind_clust, 1, function(x) sum(x==0)))]
    families_clust <- colnames(null_disp_ind_sub)[null_disp_ind_sub[worst_gene,] == 1]
    cat(paste("Dispersion estimation for cluster", i_clust, "the number of families is", length(families_clust),", number of genes is ", length(genes_clust)," \n"))
    counts_clust <- matrix(NA, nrow = length(genes_clust), ncol = 3 * length(families_clust))
    for(n_hybrid in 1:length(families_clust)){
      counts_clust[,(n_hybrid - 1) * 3 + 1] <- raw_use[genes_clust, hybrid_info[hybrid_info$hybrid == families_clust[n_hybrid],"female"]]
      counts_clust[,(n_hybrid - 1) * 3 + 2] <- raw_use[genes_clust, hybrid_info[hybrid_info$hybrid == families_clust[n_hybrid],"male"]]
      counts_clust[,(n_hybrid - 1) * 3 + 3] <- raw_use[genes_clust, families_clust[n_hybrid]]
    }
    group_clust <- rep(paste("G", 1:length(families_clust), sep = ""), each = 3)
    DISP[genes_clust] <- Disp_est_null_family(counts = counts_clust, group = group_clust)
  }
  cat("Dispersion estimation summary:\n")
  print(summary(DISP))
  # #SS2: fit a loess curve to predict the dispersions for genes whose number of all-null families do not reach the minimal requirement all-null families
    loess_x <- apply(MP, 1, median)
    x <- loess_x[!is.na(DISP)]
    y <- DISP[!is.na(DISP)]
    # cm_G <- intersect(names(y), rownames(dge$counts))
    # df <- data.frame(
    #   mine = log10(y[cm_G]),
    #   theirs = log10(dge$tagwise.dispersion[rownames(dge$counts) %in% names(y)])
    # )
    # ggplot(df) +
    #   geom_point(aes(x=mine, y=theirs), size = 0.5, alpha =0.2)+geom_abline(slope = 1, intercept = 0)+
    #   annotate(
    #     "text",
    #     x = -Inf, y = Inf, hjust = -0.05, vjust = 1.1,
    #     label = sprintf("CCC = %.3f\nSpearman = %.3f", DescTools::CCC(df$mine, df$theirs)$rho.c$est, cor(df$mine, df$theirs, method = "spearman"))
    #   )
    if(sd(x) > 0){
      new_x <- loess_x[is.na(DISP)]
      loess_fit <- loess(log(y)~x, span = 0.01)
      pred_disp <- exp(predict(loess_fit, newdata = new_x))
      extra <- new_x[is.na(pred_disp) | pred_disp <= 0]
      intra <- new_x[!is.na(pred_disp) & pred_disp > 0 ]
      intra_disp <- pred_disp[!is.na(pred_disp) & pred_disp > 0 ]
      extra_disp <- ifelse(extra < min(x), y[which.min(x)], y[which.max(x)])
      pred_disp[is.na(pred_disp) | pred_disp <= 0] <- extra_disp
      DISP[is.na(DISP)] <- pred_disp
      }
    dispersion_mat <- matrix(DISP, nrow = nrow(raw_use), ncol = nrow(hybrid_info))
    colnames(dispersion_mat) <- hybrid_info$hybrid
    rownames(dispersion_mat) <- rownames(raw_use)
    # cm_G <- intersect(names(c(y, pred_disp[!is.na(pred_disp)])), rownames(dge$counts))
    # df <- data.frame(
    #   mine = log10(c(y,pred_disp)[cm_G]),
    #   theirs = log10(dge$tagwise.dispersion[rownames(dge$counts) %in% cm_G])
    # )
    # ggplot(df) +
    #   geom_point(aes(x=mine, y=theirs), size = 0.5, alpha =0.2)+geom_abline(slope = 1, intercept = 0)+
    #   annotate(
    #     "text",
    #     x = -Inf, y = Inf, hjust = -0.05, vjust = 1.1,
    #     label = sprintf("CCC = %.3f\nSpearman = %.3f", DescTools::CCC(df$mine, df$theirs)$rho.c$est, cor(df$mine, df$theirs, method = "spearman"))
    #   )
##### Step 3: NB estimation with known dispersion matrix ####
  NB_gene_func <- function(gene_ind){
    gene <- raw_use[gene_ind,]
    gene_name <- rownames(raw_use)[gene_ind]
    pval <- c()
    if(sd(gene) == 0){pval <- rep(1, nrow(hybrid_info))}
    for(hb_num in 1:nrow(hybrid_info)){
      female <- as.numeric(gene[hybrid_info$female[hb_num]])
      male <- as.numeric(gene[hybrid_info$male[hb_num]])
      hybrid <- as.numeric(gene[hybrid_info$hybrid[hb_num]])
      dispersion <- dispersion_mat[gene_name, hybrid_info$hybrid[hb_num]]
      if(dispersion == 0){pval[hb_num] <- Hetero_Pois_LRT(y_g1 = female, y_g2 = male, y_g3 = hybrid,
                                                          C_1 = norm_factors_use[hybrid_info$female[hb_num]],
                                                          C_2 = norm_factors_use[hybrid_info$male[hb_num]],
                                                          C_3 = norm_factors_use[hybrid_info$hybrid[hb_num]])}
      else{pval[hb_num] <- Hetero_NB_LRT(y_g1 = female, y_g2 = male, y_g3 = hybrid,
                                         C_1 = norm_factors_use[hybrid_info$female[hb_num]],
                                         C_2 = norm_factors_use[hybrid_info$male[hb_num]],
                                         C_3 = norm_factors_use[hybrid_info$hybrid[hb_num]],
                                         dispersion = dispersion)}
      names(pval)[hb_num] <- hybrid_info$hybrid[hb_num]
    }
    return(pval)
  }
  
  pval_NB <- do.call(rbind, lapply(1:nrow(raw_use),  NB_gene_func))
  rownames(pval_NB) <- rownames(raw_use)
  qval_NB <- apply(pval_NB, 2, function(x){p.adjust(x, method = "fdr")})
```





## 5. Test MPH with replicates


```{r}
rep_reps <- rownames(metadata)[metadata$Genotype %in% meta_rep$Genotype]
counts_rep_rep <- counts[, rep_reps]
meta_rep_rep <- droplevels(metadata[rep_reps,])
```


```{r}
# ==============================================================================
# PART 1: PREPROCESSING AND ESTIMATION OF DISPERSION
# ==============================================================================

# 1. Create DGEList
dge <- DGEList(counts = counts_rep_rep)

# 2. Normalization (TMM)
dge <- calcNormFactors(dge)

# 3. Filtering
# MATRIX 1: Just Biology. We want to keep genes expressed in any Genotype.
design_filter <- model.matrix(~ Genotype, data = meta_rep_rep)
keep <- filterByExpr(dge, design_filter)
dge <- dge[keep, , keep.lib.sizes=FALSE]

# 4. Dispersion Estimation
# MATRIX 2: Biology + Technical. We need to subtract Batch noise to find real Dispersion.
# Note: Treat Batch as Fixed here to help edgeR "clean" the variance.
design_disp <- model.matrix(~ Genotype + REP.BAT, data = meta_rep_rep)
# design <- model.matrix(
#   ~ 0 + as.factor(sampleinfo$Genotype) + as.factor(sampleinfo$REP.BAT) # REP.BAT for batch nested within replicate
# ) this is from the author's code
dge <- estimateDisp(dge, design_disp)

# 5. Extract the "Clean" Dispersion for glmmTMB
# edgeR gives 'dispersion' (phi). glmmTMB wants 'theta' (size).
# Relationship: theta = 1 / phi
gene_thetas <- 1 / dge$tagwise.dispersion
names(gene_thetas) <- rownames(dge$counts)


# ==============================================================================
# PART 2: BUILD THE MASTER CONTRAST LIST (Run Once)
# ==============================================================================

# Get all genotype levels exactly as they appear in the factor
all_levs <- levels(meta_rep_rep$Genotype)

# Identify Hybrid names (assuming they contain underscores or specific pattern)
# Adjust pattern if needed. Here I assume Hybrids have "_"
hybrid_names <- grep("_", all_levs, value = TRUE) 

# Generate the list of numeric vectors
master_contrast_list <- setNames(lapply(hybrid_names, function(h) {
  
  # Logic: "B73_ M0001" -> c("B73", "M0001")
  # Uses trimws() to handle spaces like " M0001"
  parents <- trimws(unlist(strsplit(h, "_")))
  
  # Check if both parents exist in the dataset
  if(!all(parents %in% all_levs)) return(NULL)
  
  # Create zero vector
  vec <- setNames(numeric(length(all_levs)), all_levs)
  
  # Set arithmetic MPH coefficients: Hybrid - 0.5*P1 - 0.5*P2
  vec[h] <- 1
  vec[parents] <- -0.5
  
  return(vec)
}), hybrid_names)

# Remove any hybrids where parents weren't found (remove NULLs)
master_contrast_list <- master_contrast_list[!sapply(master_contrast_list, is.null)]

print(paste("Created contrasts for", length(master_contrast_list), "hybrids."))

# ==============================================================================
# PART 3: THE GENE WORKER FUNCTION
# ==============================================================================

run_all_families_test <- function(count_vec, theta_fixed, metadata, contrast_list) {
  
  df_tmp <- data.frame(Count = as.numeric(count_vec) + 1, # This is to avoid convergence problem in glmmTMB 
                       Genotype = metadata$Genotype,
                       REP.BAT = metadata$REP.BAT,
                       REP.BAT.SYS = metadata$REP.BAT.SYS)
  
  # A. Try Full Model (System Random Effect)
  fit <- tryCatch({
    glmmTMB(Count ~ Genotype + REP.BAT + (1|REP.BAT.SYS), 
            data = df_tmp, 
            family = nbinom2,
            start = list(betadisp = log(theta_fixed)), 
            map = list(betadisp = factor(NA)),
            control = glmmTMBControl(optCtrl=list(iter.max=1000, eval.max=1000))
    )
  }, error = function(e) return(NULL))
  
  # B. Fallback: If Full Model fails (Singular fit / Convergence), try Simple Model
  if(is.null(fit)) {
    fit <- tryCatch({
      glmmTMB(Count ~ Genotype + REP.BAT, 
              data = df_tmp, 
              family = nbinom2,
              start = list(betadisp = log(theta_fixed)), 
              map = list(betadisp = factor(NA))
      )
    }, error = function(e) return(NULL))
  }
  
  # C. If both fail, return NAs
  if(is.null(fit)) return(rep(NA, length(contrast_list)))
  
  # D. Calculate Stats
  # Wrap in tryCatch because emmeans can fail on very weird fits
  pvals <- tryCatch({
    emm <- emmeans(fit, ~ Genotype)
    emm_arith <- regrid(emm) # Transform to arithmetic scale
    
    # Run ALL contrasts simultaneously
    res <- contrast(emm_arith, method = contrast_list)
    
    # Return just the p-values as a named vector
    s <- summary(res)
    stats <- s$p.value
    names(stats) <- s$contrast
    return(stats)
    
  }, error = function(e) return(rep(NA, length(contrast_list))))
  
  return(pvals)
}

# ==============================================================================
# PART 4: RUN THE LOOP (Parallel) AND FORMAT OUTPUT
# ==============================================================================


# 1. Detect your total cores
total_cores <- detectCores()

# 2. Decide how many to use
# Best Practice: Use (Total - 1) or (Total - 2) so your computer doesn't freeze.
# If you have 8 cores, use 6 or 7.
use_cores <- max(1, total_cores - 2)

print(paste("Your system has", total_cores, "cores."))
print(paste("Using", use_cores, "cores for processing."))

# Extract the clean count matrix
count_matrix <- dge$counts

# Run in parallel (Adjust mc.cores based on your CPU)
# Use mc.cores = 1 if on Windows
results_replicates <- mclapply(seq_len(nrow(count_matrix)), function(i) {
  
  gene_id <- rownames(count_matrix)[i]
  theta   <- gene_thetas[gene_id]
  counts  <- count_matrix[i, ]
  
  # Run function
  run_all_families_test(counts, theta, meta_rep_rep, master_contrast_list)
  
}, mc.cores = use_cores) 

# ==============================================================================
# PART 5: CONVERT LIST TO MATRIX
# ==============================================================================

# Bind rows together
pval_replicates <- do.call(rbind, results_replicates)

# Assign Gene Names to Rows
rownames(pval_replicates) <- rownames(dge$counts)

# Check dimensions
print(dim(pval_replicates))
# Should be: [Number of Genes] x [Number of Hybrids]

# Preview
head(pval_replicates)
```

## 6. Results comparison 


```{r}
generate_deep_diagnosis <- function(target_family, 
                                    counts_matrix, 
                                    metadata, 
                                    hybrid_info, 
                                    disp_dat_mine,   
                                    disp_dat_theirs,
                                    fdr_mine_full,    
                                    fdr_theirs_full,
                                    common_genes,
                                    alpha = 0.05) {
  
  # --- 1. SETUP: FIND SAMPLE IDs ---
  parents <- trimws(unlist(strsplit(target_family, "_")))
  
  # Robust Lookup: Get the column name passed in the 1-col matrix
  sample_name_in_matrix <- colnames(fdr_mine_full)
  match_idx <- which(hybrid_info$hybrid == sample_name_in_matrix)
  
  if(length(match_idx) == 0) stop(paste("Sample", sample_name_in_matrix, "not found in hybrid_info"))
  
  my_hybrid_id <- hybrid_info$hybrid[match_idx]
  my_female_id <- hybrid_info$female[match_idx]
  my_male_id   <- hybrid_info$male[match_idx]
  
  # Find Their Replicates
  all_hybrid_reps <- rownames(metadata)[metadata$Genotype == target_family | metadata$Genotype == gsub("_", "x", target_family)]
  all_p1_reps     <- rownames(metadata)[metadata$Genotype == parents[1]]
  all_p2_reps     <- rownames(metadata)[metadata$Genotype == parents[2]]
  
  # --- 2. SUBSET DATA ---
  # Ensure we only work on the intersected genes passed to the function
  mat <- counts_matrix[common_genes, c(all_hybrid_reps, all_p1_reps, all_p2_reps)]
  
  # --- 3. CALCULATE METRICS ---
  
  # Mine (Single Rep)
  my_val_h   <- mat[common_genes, my_hybrid_id]
  my_val_f   <- mat[common_genes, my_female_id]
  my_val_m   <- mat[common_genes, my_male_id]
  my_val_min <- pmin(my_val_h, my_val_f, my_val_m)
  my_val_mp  <- (my_val_f + my_val_m) / 2
  
  my_mph_raw <- my_val_h - my_val_mp
  my_mph     <- (my_val_h - my_val_mp) / (my_val_mp) 
  
  my_fdr     <- fdr_mine_full[common_genes, 1] # Take the data from the single column
  my_disp    <- disp_dat_mine[common_genes, "Dispersion"]
  
  # Theirs (Benchmark Means)
  their_mean_h  <- rowMeans(mat[common_genes, all_hybrid_reps, drop=FALSE])
  their_mean_f  <- rowMeans(mat[common_genes, all_p1_reps, drop=FALSE]) 
  their_mean_m  <- rowMeans(mat[common_genes, all_p2_reps, drop=FALSE]) 
  their_mean_mp <- (their_mean_f + their_mean_m) / 2
  
  their_mph_raw <- their_mean_h - their_mean_mp
  their_mph     <- (their_mean_h - their_mean_mp) / (their_mean_mp)
  
  their_fdr     <- fdr_theirs_full[common_genes, target_family]
  their_disp    <- disp_dat_theirs[common_genes, "Dispersion"]
  
  # Noise Check (Max CV of family)
  calc_cv <- function(sub_mat) apply(sub_mat, 1, function(x) sd(x)/mean(x))
  max_cv <- pmax(calc_cv(mat[common_genes, all_hybrid_reps, drop=FALSE]), 
                 calc_cv(mat[common_genes, all_p1_reps, drop=FALSE]), 
                 calc_cv(mat[common_genes, all_p2_reps, drop=FALSE]), na.rm=TRUE)
  
  # Deviations
  dev_h <- abs(my_val_h - their_mean_h) / (their_mean_h + 1)
  dev_f <- abs(my_val_f - their_mean_f) / (their_mean_f + 1)
  dev_m <- abs(my_val_m - their_mean_m) / (their_mean_m + 1)
  
  # --- 4. CLASSIFICATIONS ---
  sig_m <- my_fdr < alpha & !is.na(my_fdr)
  sig_t <- their_fdr < alpha & !is.na(their_fdr)
  
  class_vec <- rep("TN", length(common_genes))
  class_vec[sig_m & sig_t]  <- "TP"
  class_vec[sig_m & !sig_t] <- "FP"
  class_vec[!sig_m & sig_t] <- "FN"
  
  # --- 5. DIAGNOSTICS FLAGS ---
  
  # Flag 1: Dispersion Underestimation (The cause of FP)
  # Logic: My estimate was < 0.2 (Stable), Reality was > 0.4 (Noisy)
  flag_disp_low <- (my_disp < 0.2) & (their_disp > 0.4)
  # other side
  flag_disp_high <- my_disp > (2 * their_disp)
  
  # Flag 2: High Noise
  flag_noise <- (max_cv > 0.5)
  
  # Flag 3: Outlier Sample
  # Logic: My Effect > 2x Their Effect 
  flag_outlier_high <- (abs(my_mph) > (2 * abs(their_mph)))
  flag_outlier_low <- abs(my_mph) < (0.5 * abs(their_mph))

  # Flag 4: Low Counts
  flag_low <- my_val_min < 10
  
  # Flag 5: Weak Signal (Practical Filter)
  flag_weak <- abs(my_mph) < 0.5 # h/mp in (0.5, 1.5)
  
  
  # Flag 5: Confident calls 
  flag_confident <- abs(their_mph) > 0.2 #(their_mph > 2) | (their_mph < -2/3)
  
  # Construct Reason String
  is_fp <- class_vec == "FP"
  reason_vec <- rep(NA, length(common_genes))
  reason_vec[is_fp] <- paste0(
    ifelse(flag_disp_low[is_fp],    "DispersionError + ", ""),
    ifelse(flag_noise[is_fp],   "HighNoise + ", ""),
    ifelse(flag_outlier_high[is_fp], "Outlier + ", ""),
    ifelse(flag_low[is_fp],     "LowCount + ", "")
  )
  reason_vec <- sub(" \\+ $", "", reason_vec)
  reason_vec[is_fp & reason_vec == ""] <- "Unknown"
  

  
  is_fn <- class_vec == "FN"
  fn_reason_vec <- rep(NA, length(common_genes))
  fn_reason_vec[is_fn] <- paste0(
    ifelse(flag_low[is_fn],         "LowCount + ", ""),
    ifelse(flag_outlier_low[is_fn], "SignalDilution + ", ""),
    ifelse(flag_disp_high[is_fn],   "DispOverestimation + ", "")
  )
  fn_reason_vec <- sub(" \\+ $", "", fn_reason_vec)
  fn_reason_vec[is_fn & fn_reason_vec == ""] <- "Unknown"
  
  # --- 6. SUBSET DEFINITIONS ---
  
  # Subset A: Quantifiable
  flag_quantifiable <- !flag_low
  
  # Subset B: Quantifiable + Strong
  flag_quantifiable_strong <- flag_quantifiable & !flag_weak
  
  # Subset C: Representative (The Gold Standard)
  # Must be Strong, Quantifiable, Accurate (Dev < 0.5), and Stable (CV < 0.5)
  flag_representative <- flag_quantifiable_strong & 
                         (dev_h < 0.5) & (dev_f < 0.5) & (dev_m < 0.5) & 
                         (max_cv < 0.5)
  
  flag_disp_correct <- !flag_disp_high & !flag_disp_low
  flag_signal_correct <- !flag_outlier_high & !flag_outlier_low

  
  # --- 7. ASSEMBLE ---
  df <- data.frame(
    Hybrid = target_family, 
    Gene = common_genes,
    
    # Subsets
    Is_Quantifiable = flag_quantifiable,
    Is_Quantifiable_Strong = flag_quantifiable_strong,
    Is_Representative = flag_representative,
    Is_Disp_Correct = flag_disp_correct,
    Is_Signal_Correct = flag_signal_correct,
    Is_Count_Close =  (dev_h < 0.5) & (dev_f < 0.5) & (dev_m < 0.5),
    Is_Rep_Confidenct = flag_confident,
    
    # Result
    Classification = class_vec,
    FP_Reason_Str = reason_vec, 
    
    # Flags
    Is_Disp_High = flag_disp_high,
    Is_Disp_Low = flag_disp_low,
    Is_High_Noise = flag_noise,
    Is_Outlier_High  = flag_outlier_high,
    Is_Outlier_Low = flag_outlier_low,
    Is_Low_Count  = flag_low,
    Is_Weak = flag_weak,
    
    
    # Data
    My_FDR = my_fdr, Their_FDR = their_fdr,
    My_MPH = my_mph, Their_MPH = their_mph,
    My_MPH_Raw = my_mph_raw, Their_MPH_Raw = their_mph_raw,
    My_Disp = my_disp, Their_Disp = their_disp,
    My_Hybrid_Count = my_val_h, Their_Hybrid_Mean = their_mean_h,
    My_Female_Count = my_val_f, Their_Female_Mean = their_mean_f,
    My_Male_Count = my_val_m, Their_Male_Mean = their_mean_m,
    My_MP = my_val_mp, Their_MP = their_mean_mp,
    
    Their_CV  = max_cv # We return the Max CV of the family
  )
  
  return(df)
}




get_performance_metrics <- function(df, stage_name) {
  
  # 1. Classification Metrics (Cutoff based)
  TP <- sum(df$Classification == "TP")
  FP <- sum(df$Classification == "FP")
  FN <- sum(df$Classification == "FN")
  
  precision <- ifelse((TP+FP)==0, 0, TP / (TP + FP))
  recall    <- ifelse((TP+FN)==0, 0, TP / (TP + FN))
  
  # Filter for valid data
  valid_cor <- df %>% filter(!is.na(My_FDR) & !is.na(Their_FDR))
  
  # 2. Ranking Metrics
  spearman  <- cor(valid_cor$My_FDR, valid_cor$Their_FDR, method = "spearman")
  
  # CCC
  log_mine   <- -log10(pmax(valid_cor$My_FDR, 1e-300))
  log_theirs <- -log10(pmax(valid_cor$Their_FDR, 1e-300))
  ccc_val    <- CCC(log_mine, log_theirs)$rho.c$est
  
  # 3. AUC CALCULATION (New)
  # Define Truth: Did the Benchmark call it Significant?
  # Truth = 1 (TP or FN), False = 0 (FP or TN)
  truth_labels <- valid_cor$Classification %in% c("TP", "FN")
  
  # Define Prediction Score: We use (1 - FDR) so high score = high confidence
  pred_scores  <- 1 - valid_cor$My_FDR
  
  auc_val <- NA
  # AUC requires both True and False cases to exist in the data
  if (length(unique(truth_labels)) == 2) {
    # Suppress messages/warnings from pROC
    roc_obj <- tryCatch({
      roc(response = truth_labels, predictor = pred_scores, quiet = TRUE, direction = "<")
    }, error = function(e) return(NULL))
    
    if(!is.null(roc_obj)) {
      auc_val <- as.numeric(roc_obj$auc)
    }
  }
  
  # 4. TOP 10000 OVERLAP
  valid_cor$UniqueID <- paste0(valid_cor$Gene, "_", valid_cor$Hybrid)
  top_n <- min(10000, nrow(valid_cor))
  
  my_top_ids <- valid_cor %>% arrange(My_FDR) %>% head(top_n) %>% pull(UniqueID)
  their_top_ids <- valid_cor %>% arrange(Their_FDR) %>% head(top_n) %>% pull(UniqueID)
  
  top_overlap <- length(intersect(my_top_ids, their_top_ids)) / top_n
  
  return(data.frame(
    Stage       = stage_name,
    N_Pairs     = nrow(df),
    Precision   = round(precision * 100, 1),
    Recall      = round(recall * 100, 1),
    AUC         = round(auc_val, 3),          
    Spearman    = round(spearman, 3),
    CCC_LogP    = round(ccc_val, 3),
    Top10000_Ovrl = round(top_overlap * 100, 1)
  ))
}
```



```{r}
# ==============================================================================
# PHASE 1: FIND THE BEST REPLICATE (Correct FDR Order)
# ==============================================================================

# 1. Start with FULL Raw P-value Matrices
raw_mine_full   <- pval_NB
raw_theirs_full <- pval_replicates

# 2. Calculate FDR on FULL Data 
fdr_mine_full   <- apply(raw_mine_full, 2, p.adjust, method = "BH")
fdr_theirs_full <- apply(raw_theirs_full, 2, p.adjust, method = "BH")

# 3. NOW Intersect Genes (Subset the FDR matrices, not the P-value matrices)
common_genes <- intersect(rownames(fdr_mine_full), rownames(fdr_theirs_full))
common_genes <- intersect(common_genes, names(y))

# Subset the FDR matrices
fdr_mine_sub   <- fdr_mine_full[common_genes, ]
fdr_theirs_sub <- fdr_theirs_full[common_genes, ]

# 4. Build Mapping Table (Same as before)
original_names <- colnames(fdr_mine_sub)
target_names   <- gsub("x", "_", sub("_.*", "", original_names)) 
valid_targets  <- gsub(" ", "", colnames(fdr_theirs_sub))
colnames(fdr_theirs_sub) <- valid_targets

mapping_df <- data.frame(Sample = original_names, Target = target_names, stringsAsFactors = F) %>%
  filter(Target %in% valid_targets)

# 5. Loop to Score Replicates (Using pre-calculated FDRs)
scores_list <- lapply(1:nrow(mapping_df), function(i) {
  samp <- mapping_df$Sample[i]
  targ <- mapping_df$Target[i]
  
  # Use the SUBSETTED FDR matrices here
  fdr_m <- fdr_mine_sub[, samp]
  fdr_t <- fdr_theirs_sub[, targ]
  
  TP <- sum(fdr_m < 0.05 & fdr_t < 0.05, na.rm=T)
  FP <- sum(fdr_m < 0.05 & fdr_t >= 0.05, na.rm=T)
  FN <- sum(fdr_m >= 0.05 & fdr_t < 0.05, na.rm=T)
  
  prec <- ifelse((TP+FP)==0, 0, TP/(TP+FP))
  rec  <- ifelse((TP+FN)==0, 0, TP/(TP+FN))
  f1   <- ifelse((prec+rec)==0, 0, 2*(prec*rec)/(prec+rec))
  
  return(data.frame(Sample=samp, Target=targ, F1=f1))
})

# 6. Pick Winners
best_reps <- do.call(rbind, scores_list) %>% group_by(Target) %>% arrange(desc(F1)) %>% slice(1) %>% ungroup()

# 7. Create CLEAN FDR Matrix for Phase 2
# We pull the specific columns from the ALREADY adjusted matrix
fdr_mine_clean <- fdr_mine_sub[, best_reps$Sample]
colnames(fdr_mine_clean) <- best_reps$Target

# Clean Benchmark Matrix
fdr_theirs_clean <- fdr_theirs_sub[, best_reps$Target]
# ==============================================================================
# 3. RUN DIAGNOSTIC LOOP
# ==============================================================================

# Prepare Dispersion Data (Ensuring alignment)
disp_dat_mine <- data.frame(Dispersion=c(y, intra_disp, extra_disp), MP=c(x, intra, extra),
                            type = factor(rep(c("Estimated","Predicted","Extrapolated"), 
                                              c(length(x), length(intra), length(extra))), 
                                          levels = c("Estimated","Predicted","Extrapolated")))
disp_dat_theirs <- data.frame(Dispersion = dge$tagwise.dispersion, Type = "Estimated", row.names = rownames(dge$counts))

disp_dat_mine <- disp_dat_mine[common_genes,]
disp_dat_theirs <- disp_dat_theirs[common_genes,]

all_diag_list <- lapply(colnames(fdr_mine_clean), function(target_family) {
  tryCatch({
    # Lookup the specific Sample ID for this Family
    my_sample_id <- best_reps$Sample[best_reps$Target == target_family]
    
    # Create 1-col matrix with Sample ID Name
    fdr_vec <- fdr_mine_clean[, target_family]
    fdr_mine_input <- matrix(fdr_vec, ncol=1, dimnames=list(names(fdr_vec), my_sample_id))
    
    generate_deep_diagnosis(
      target_family   = target_family,
      counts_matrix   = dge$counts[common_genes,]+1,
      metadata        = metadata,
      hybrid_info     = hybrid_info,
      disp_dat_mine   = disp_dat_mine,
      disp_dat_theirs = disp_dat_theirs, 
      fdr_mine_full   = fdr_mine_input, 
      fdr_theirs_full = fdr_theirs_clean,
      common_genes    = common_genes,
      alpha = 0.1
    )
  }, error = function(e) { warning(e); return(NULL) })
})

master_diagnosis_df <- do.call(rbind, all_diag_list[!sapply(all_diag_list, is.null)])

# ==============================================================================
# 4. REPORT GENERATION
# ==============================================================================

# Generate Rows
r1 <- get_performance_metrics(master_diagnosis_df, "1. Raw Output")
r2 <- get_performance_metrics(subset(master_diagnosis_df, Is_Quantifiable), "2. Remove low count (min<10)")
r3 <- get_performance_metrics(subset(master_diagnosis_df, Is_Quantifiable_Strong), "3. Remove low count and weak signal(h=(0.5, 1.5)mp) ")
r4 <- get_performance_metrics(subset(master_diagnosis_df, Is_Representative), "4. Remove low count and weak signal and noisy(cv>0.5)")
r5 <- get_performance_metrics(subset(master_diagnosis_df, Is_Disp_Correct), "5. Remove Disp Error")
r6 <- get_performance_metrics(subset(master_diagnosis_df, Is_Signal_Correct), "6. Remove Outliers")
r7 <- get_performance_metrics(subset(master_diagnosis_df, Is_Count_Close), "7. Remove Not Similar Counts")


final_table <- rbind(r1, r2, r3, r4, r5, r6, r7)
final_table$Data_Remaining_Pct <- round(final_table$N_Pairs / nrow(master_diagnosis_df) * 100, 1)

print(final_table)
```

```{r}
r1 <- get_performance_metrics(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), "1. Raw Output")
r2 <- get_performance_metrics(subset(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), Is_Quantifiable), "2. Remove low count (min<10)")
r3 <- get_performance_metrics(subset(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), Is_Quantifiable_Strong), "3. Remove low count and weak signal(absmph<0.3) ")
r4 <- get_performance_metrics(subset(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), Is_Representative), "4. Remove low count and weak signal and noisy(cv>0.5)")
r5 <- get_performance_metrics(subset(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), Is_Disp_Correct), "5. Remove Disp Error")
r6 <- get_performance_metrics(subset(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), Is_Signal_Correct), "6. Remove Outliers")
r7 <- get_performance_metrics(subset(master_diagnosis_df %>% subset(Is_Rep_Confidenct == TRUE), Is_Count_Close), "7. Remove Not Similar Counts")


final_table <- rbind(r1, r2, r3, r4, r5, r6, r7)
final_table$Data_Remaining_Pct <- round(final_table$N_Pairs / nrow(master_diagnosis_df) * 100, 1)

print(final_table)
```


# Comparison using in house replicated dataset

We drop sample B18.2nd, and B21.2nd to make sure we have two replicates per genotype


```{r}

logLik_NB <- function(y, C, mu, r) { sum(y * log(C * mu) - (y + r) * log(C * mu + r)) }
score_NB  <- function(y, C, mu, r) { sum( (y / mu) - ((y + r) * C) / (C * mu + r) ) }
hessian_NB<- function(y, C, mu, r) { sum( -(y / mu^2) + ((y + r) * C^2) / (C * mu + r)^2 ) }

solve_H1_1D <- function(y, C, r) {
  mu <- sum(y) / sum(C)
  if(is.nan(mu) || mu == 0) return(1e-4)
  for(k in 1:25) {
    S <- score_NB(y, C, mu, r); H <- hessian_NB(y, C, mu, r)
    step <- S / H
    mu_new <- mu - step
    iter_half <- 0
    while(mu_new <= 0 && iter_half < 10) { step <- step * 0.5; mu_new <- mu - step; iter_half <- iter_half + 1 }
    if(mu_new <= 0) mu_new <- 1e-5
    if(abs(mu_new - mu) < 1e-6 * mu) return(mu_new)
    mu <- mu_new
  }
  return(mu)
}

test_Family_LRT <- function(y_list, C_list, phi) {
  r <- 1/phi
  mu_hat <- numeric(3); logL_H1 <- 0
  for(i in 1:3) {
    mu_hat[i] <- solve_H1_1D(y_list[[i]], C_list[[i]], r)
    logL_H1 <- logL_H1 + logLik_NB(y_list[[i]], C_list[[i]], mu_hat[i], r)
  }

  theta <- c(mu_hat[1], mu_hat[2])
  for(iter in 1:25) {
    m1 <- theta[1]; m2 <- theta[2]; m3 <- (m1 + m2)/2
    d1 <- c(score_NB(y_list[[1]],C_list[[1]],m1,r), hessian_NB(y_list[[1]],C_list[[1]],m1,r))
    d2 <- c(score_NB(y_list[[2]],C_list[[2]],m2,r), hessian_NB(y_list[[2]],C_list[[2]],m2,r))
    d3 <- c(score_NB(y_list[[3]],C_list[[3]],m3,r), hessian_NB(y_list[[3]],C_list[[3]],m3,r))

    grad <- c(d1[1] + 0.5*d3[1], d2[1] + 0.5*d3[1])
    H11 <- d1[2] + 0.25*d3[2]; H22 <- d2[2] + 0.25*d3[2]; H12 <- 0.25*d3[2]

    det <- H11*H22 - H12*H12
    if(abs(det) < 1e-10) break
    delta1 <- (H22*grad[1] - H12*grad[2])/det
    delta2 <- (-H12*grad[1] + H11*grad[2])/det

    theta_new <- theta - c(delta1, delta2)
    step_size <- 1
    while(any(theta_new <= 0) && step_size > 1e-4) {
      step_size <- step_size * 0.5
      theta_new <- theta - (c(delta1, delta2) * step_size)
    }
    if(max(abs(theta_new - theta)) < 1e-5) { theta <- theta_new; break }
    theta <- theta_new
  }
  logL_H0 <- logLik_NB(y_list[[1]],C_list[[1]],theta[1],r) +
    logLik_NB(y_list[[2]],C_list[[2]],theta[2],r) +
    logLik_NB(y_list[[3]],C_list[[3]],(theta[1]+theta[2])/2,r)

  LRT <- 2 * (logL_H1 - logL_H0)
  if(LRT < 0 || is.na(LRT)) LRT <- 0
  return(pchisq(LRT, df=1, lower.tail=FALSE))
}
```

## 1. Replicated analysis
```{r}
samples2exclude <- c("B18.2nd", "B21.2nd")
data_rep <- read.table("small_family_size/allSamples.featurecounts.add2nd.txt", sep = "\t", header = TRUE, row.names = "Geneid")
# load("../IntermediateRes/WithNormFactors/Normfac_NBpqvalue_19880.RData")
group_info <- read.table("small_family_size/group_info.txt", sep = "\t", header=TRUE)
group_info <- group_info %>%
  mutate(Genotype = gsub("X", "/", Condition)) %>%
  subset(!(ID %in% samples2exclude))%>%
  { 
    rownames(.) <- .$ID
    .
  }

hybrid_info <- group_info %>%
  filter(grepl("/", Genotype)) %>%
  mutate(hybrid = Genotype) %>%
  separate(
    col   = Genotype,
    into  = c("female", "male"),
    sep   = "/",
    remove = TRUE
  ) %>%
  select(hybrid, female, male) %>%
  distinct() %>%
  { 
    rownames(.) <- NULL
    .
  }



counts <- data_rep[,group_info$ID]

Big_C <- DESeq2::estimateSizeFactorsForMatrix(counts); if(any(is.na(Big_C))) Big_C[] <- 1

# valid_genes <- intersect(rownames(pval_NB),rownames(counts))

valid_genes <- rownames(counts)[apply(counts, 1, function(x){sum(x==0)==0})]

Big_Counts <- counts[valid_genes,] 
Big_Counts_Norm <- sweep(Big_Counts, 2, Big_C[colnames(Big_Counts)], "/")
Big_Group <- factor(group_info$Genotype)
y_dge <- edgeR::DGEList(counts=Big_Counts, group=Big_Group)
y_dge$offset <- matrix(log(Big_C), nrow=nrow(Big_Counts), ncol=ncol(Big_Counts), byrow=TRUE)
y_dge <- edgeR::estimateDisp(y_dge, model.matrix(~0+Big_Group), robust=TRUE)
global_phi <- y_dge$tagwise.dispersion




n_genes <- nrow(Big_Counts)
n_families <- nrow(hybrid_info)

p_mat_rep <- matrix(NA, nrow=n_genes, ncol=n_families)
colnames(p_mat_rep) <- hybrid_info$hybrid
rownames(p_mat_rep) <- rownames(Big_Counts)

for (fam in unique(hybrid_info$hybrid)) {
  
  # Get parent names
  hi_fam <- hybrid_info[hybrid_info$hybrid == fam, , drop = FALSE]
  female_geno <- hi_fam$female[1]
  male_geno   <- hi_fam$male[1]

  # Identify column names for this family
  H_ids <- group_info$ID[group_info$Genotype == fam]
  F_ids <- group_info$ID[group_info$Genotype == female_geno]
  M_ids <- group_info$ID[group_info$Genotype == male_geno]
  
  # Skip if missing parents (Safety check)
  if(length(F_ids) == 0 || length(M_ids) == 0) next 
      
  # Extract Data
  y_f <- Big_Counts[, F_ids, drop=FALSE]
  y_m <- Big_Counts[, M_ids, drop=FALSE]
  y_h <- Big_Counts[, H_ids, drop=FALSE]
  
  c_f <- Big_C[F_ids]
  c_m <- Big_C[M_ids]
  c_h <- Big_C[H_ids]
      
  # Run Newton-Raphson LRT gene-by-gene
  fam_pvals <- numeric(n_genes)
  
  # NOTE: Ensure test_Family_LRT is loaded in your environment
  for(g in 1:n_genes) {
      fam_pvals[g] <- test_Family_LRT(list(y_f[g,], y_m[g,], y_h[g,]), 
                                      list(c_f, c_m, c_h), 
                                      global_phi[g])
  }
  
  p_mat_rep[, fam] <- fam_pvals
}

# Adjust FDR globally
q_mat_rep <- apply(p_mat_rep, 2, function(x){p.adjust(x, method = "fdr")})
```


## 2. Apply two stage LRT to a single replicate 

```{r perform 2sLRT on real data, eval = FALSE}
twoStageLRT_app <- function(hybrid_info, raw_use, norm_use, norm_factors_use, gene_groups_no = 3, n_fam_thres = 2, alpha0 = 0.01){
  ##### Step 1: Gene-wise pre-test - Poisson LRT for null-family ####
  gene_test_func <- function(gene){
  # compute pvalues from test using raw counts and norm factors
    pval_family <- c()  # for Poisson LRT equal family
    if(sd(gene) == 0){
      # if this gene is equally expressed in all varieties
      pval_family <- rep(1, nrow(hybrid_info))
    }
    for(hb_num in 1:nrow(hybrid_info)){
      female <- as.numeric(gene[hybrid_info$female[hb_num]])
      male <- as.numeric(gene[hybrid_info$male[hb_num]])
      hybrid <- as.numeric(gene[hybrid_info$hybrid[hb_num]])
      
      pval_family[hb_num] <- Pois_3mean_LRT(y_g1 = female, y_g2 = male, y_g3 = hybrid,
                                            C_1 = norm_factors_use[hybrid_info$female[hb_num]],
                                            C_2 = norm_factors_use[hybrid_info$male[hb_num]],
                                            C_3 = norm_factors_use[hybrid_info$hybrid[hb_num]])
      names(pval_family)[hb_num] <- hybrid_info$hybrid[hb_num]
    }
    return(list(pval_family = pval_family))
}

  gene_est_func <- function(gene){
  # get MP, MPH ratio and diff MPH ratio from normalized data, i.e. estimation of mu_gi under global/Hg1
    MP <- c()           # mid parents
    hyb_ratio <- c()    # MPH ratio = hyb/MP
    diff_ratio <- c()   # MPH diff ratio = (hyb-MP) / MP
    family_mean <- c()  # family mean 
    if(all(gene == 0)){
      # if this gene is not expressed in all varieties
      MP <- hyb_ratio <- diff_ratio <- family_mean <- rep(0, nrow(hybrid_info))
    }
    for(hb_num in 1:nrow(hybrid_info)){
      female <- as.numeric(gene[hybrid_info$female[hb_num]])
      male <- as.numeric(gene[hybrid_info$male[hb_num]])
      hybrid <- as.numeric(gene[hybrid_info$hybrid[hb_num]])
      
      MP[hb_num] <- mean(c(female, male))
      family_mean[hb_num] <- mean(c(female, male, hybrid))
      hyb_ratio[hb_num] <- hybrid/(MP[hb_num])
      diff_ratio[hb_num] <- (hybrid - MP[hb_num]) / (MP[hb_num])
      names(MP)[hb_num] <- names(family_mean) <- names(hyb_ratio)[hb_num] <- names(diff_ratio)[hb_num] <- hybrid_info$hybrid[hb_num]
    }
    return(list(MP = MP, family_mean = family_mean, hyb_ratio = hyb_ratio, diff_ratio = diff_ratio))
  }
  
  test_result_list <- lapply(1:nrow(raw_use), function(i) gene_test_func(raw_use[i, ]))
  est_result_list <- lapply(1:nrow(norm_use), function(i) gene_est_func(norm_use[i, ]))

  
  # combine genewise pre-test as a matrix, the order of results is consistent of hybrids in hybrid_info.
  pval_family <- do.call(rbind, lapply(test_result_list, function(x) x$pval_family))
  MP <- do.call(rbind, lapply(est_result_list, function(x) x$MP))
  family_mean <- do.call(rbind, lapply(est_result_list, function(x) x$family_mean))
  diff_ratio <- do.call(rbind, lapply(est_result_list, function(x) x$diff_ratio))
  hyb_ratio <- do.call(rbind, lapply(est_result_list, function(x) x$hyb_ratio))
  rownames(pval_family) <- rownames(MP) <- rownames(hyb_ratio) <- rownames(diff_ratio) <- rownames(family_mean) <- rownames(raw_use)
  colnames(pval_family) <- colnames(MP) <- colnames(hyb_ratio) <- colnames(diff_ratio) <- colnames(family_mean) <- hybrid_info$hybrid

  qval_family <- apply(pval_family, 2, function(x){p.adjust(x, method = "fdr")})
  

##### Step 2: Estimate dispersion matrix ####
  null_disp_ind <- qval_family > alpha0
  DISP <- rep(NA, nrow(null_disp_ind)) # each gene will only have one dispersion estimation
  names(DISP) <- rownames(null_disp_ind)
  null_disp_ind_sub <- null_disp_ind[apply(null_disp_ind, 1, function(x) sum(x == 1)) >= n_fam_thres,]
  jacc_dist <- dist(null_disp_ind_sub, "binary")
  jacc_hc <- hclust(jacc_dist)
  jacc_clust <- cutree(jacc_hc, gene_groups_no)
  cat("Gene clusters size: \n")
  print(table(jacc_clust))
  for (i_clust in unique(jacc_clust)) {
    genes_clust <- rownames(null_disp_ind_sub)[jacc_clust == i_clust]
    ind_clust <- null_disp_ind_sub[genes_clust,]
    worst_gene <- genes_clust[which.max(apply(ind_clust, 1, function(x) sum(x==0)))]
    families_clust <- colnames(null_disp_ind_sub)[null_disp_ind_sub[worst_gene,] == 1]
    cat(paste("Dispersion estimation for cluster", i_clust, "the number of families is", length(families_clust),", number of genes is ", length(genes_clust)," \n"))
    counts_clust <- matrix(NA, nrow = length(genes_clust), ncol = 3 * length(families_clust))
    for(n_hybrid in 1:length(families_clust)){
      counts_clust[,(n_hybrid - 1) * 3 + 1] <- raw_use[genes_clust, hybrid_info[hybrid_info$hybrid == families_clust[n_hybrid],"female"]]
      counts_clust[,(n_hybrid - 1) * 3 + 2] <- raw_use[genes_clust, hybrid_info[hybrid_info$hybrid == families_clust[n_hybrid],"male"]]
      counts_clust[,(n_hybrid - 1) * 3 + 3] <- raw_use[genes_clust, families_clust[n_hybrid]]
    }
    group_clust <- rep(paste("G", 1:length(families_clust), sep = ""), each = 3)
    DISP[genes_clust] <- Disp_est_null_family(counts = counts_clust, group = group_clust)
  }
  cat("Dispersion estimation summary:\n")
  print(summary(DISP))
  # #SS2: fit a loess curve to predict the dispersions for genes whose number of all-null families do not reach the minimal requirement all-null families
    loess_x <- apply(MP, 1, median)
    x <- loess_x[!is.na(DISP)]
    y <- DISP[!is.na(DISP)]
    if(sd(x) > 0){
      new_x <- loess_x[is.na(DISP)]
      loess_fit <- loess(log(y)~x, span = 0.01)
      pred_disp <- exp(predict(loess_fit, newdata = new_x))
      extra <- new_x[is.na(pred_disp) | pred_disp <= 0]
      intra <- new_x[!is.na(pred_disp) & pred_disp > 0 ]
      intra_disp <- pred_disp[!is.na(pred_disp) & pred_disp > 0 ]
      extra_disp <- ifelse(extra < min(x), y[which.min(x)], y[which.max(x)])
      pred_disp[is.na(pred_disp) | pred_disp <= 0] <- extra_disp
      DISP[is.na(DISP)] <- pred_disp
      }
    dispersion_mat <- matrix(DISP, nrow = nrow(raw_use), ncol = nrow(hybrid_info))
    colnames(dispersion_mat) <- hybrid_info$hybrid
    rownames(dispersion_mat) <- rownames(raw_use)
##### Step 3: NB estimation with known dispersion matrix ####
  NB_gene_func <- function(gene_ind){
    gene <- raw_use[gene_ind,]
    gene_name <- rownames(raw_use)[gene_ind]
    pval <- c()
    if(sd(gene) == 0){pval <- rep(1, nrow(hybrid_info))}
    for(hb_num in 1:nrow(hybrid_info)){
      female <- as.numeric(gene[hybrid_info$female[hb_num]])
      male <- as.numeric(gene[hybrid_info$male[hb_num]])
      hybrid <- as.numeric(gene[hybrid_info$hybrid[hb_num]])
      dispersion <- dispersion_mat[gene_name, hybrid_info$hybrid[hb_num]]
      if(dispersion == 0){pval[hb_num] <- Hetero_Pois_LRT(y_g1 = female, y_g2 = male, y_g3 = hybrid,
                                                          C_1 = norm_factors_use[hybrid_info$female[hb_num]],
                                                          C_2 = norm_factors_use[hybrid_info$male[hb_num]],
                                                          C_3 = norm_factors_use[hybrid_info$hybrid[hb_num]])}
      else{pval[hb_num] <- Hetero_NB_LRT(y_g1 = female, y_g2 = male, y_g3 = hybrid,
                                         C_1 = norm_factors_use[hybrid_info$female[hb_num]],
                                         C_2 = norm_factors_use[hybrid_info$male[hb_num]],
                                         C_3 = norm_factors_use[hybrid_info$hybrid[hb_num]],
                                         dispersion = dispersion)}
      names(pval)[hb_num] <- hybrid_info$hybrid[hb_num]
    }
    return(pval)
  }
  
  pval_NB <- do.call(rbind, lapply(1:nrow(raw_use),  NB_gene_func))
  rownames(pval_NB) <- rownames(raw_use)
  qval_NB <- apply(pval_NB, 2, function(x){p.adjust(x, method = "fdr")})
  return(list(pval_NB = pval_NB, qval_NB = qval_NB))
}
evaluate_performance <- function(pred_p, pred_q, truth_p, truth_q, label, significance = 0.05) {
  # 1. Flatten vectors
  pv_pred <- as.vector(pred_p)
  pv_rep  <- as.vector(truth_p)
  q_pred  <- as.vector(pred_q)
  q_rep   <- as.vector(truth_q)
  
  # 2. Filter Finite values
  keep <- is.finite(pv_pred) & is.finite(pv_rep) & 
          is.finite(q_pred)  & is.finite(q_rep)
  
  pv_pred <- pv_pred[keep]; pv_rep <- pv_rep[keep]
  q_pred  <- q_pred[keep];  q_rep  <- q_rep[keep]
  
  # 3. Define Truth (Significance Threshold < 0.05)
  is_true   <- q_rep  < 0.05
  is_called <- q_pred < 0.05
  
  # 4. Counts
  TP <- sum(is_called & is_true)
  FP <- sum(is_called & !is_true)
  FN <- sum(!is_called & is_true)
  n_true <- sum(is_true)
  
  # 5. Metrics
  precision <- ifelse((TP + FP) == 0, NA, TP / (TP + FP))
  recall    <- ifelse((TP + FN) == 0, NA, TP / (TP + FN))
  
  # 6. AUC (Direction: q-values, so smaller = positive, thus direction = ">")
  auc_full <- NA
  auc_part <- NA
  
  if (length(unique(is_true)) == 2) {
    roc_obj <- roc(response = is_true, predictor = q_pred, direction = ">", quiet = TRUE)
    auc_full <- as.numeric(auc(roc_obj))
    # pAUC (FPR < 0.2)
    auc_part <- as.numeric(auc(roc_obj, partial.auc = c(1, 0.8), 
                               partial.auc.focus = "specificity", 
                               partial.auc.correct = TRUE))
  }
  
  # 7. CCC (on -log10 p-values)
  # Use pmax to avoid log(0) = -Inf
  log_pred <- -log10(pmax(pv_pred, .Machine$double.xmin))
  log_rep  <- -log10(pmax(pv_rep, .Machine$double.xmin))
  ccc_res  <- CCC(log_pred, log_rep)
  
  # 8. Top 1000 Overlap
  k <- min(1000, length(q_pred))
  top_overlap <- length(intersect(order(q_pred)[1:k], order(q_rep)[1:k]))
  
  # Return one-row data frame
  return(data.frame(
    Replicate_ID    = label,
    True_Positives  = TP,
    False_Positives = FP,
    Precision       = precision,
    Recall          = recall,
    AUC             = auc_full,
    pAUC_0.2        = auc_part,
    CCC_LogP        = ccc_res$rho.c$est,
    Top1k_Overlap   = top_overlap,
    stringsAsFactors = FALSE
  ))
}
```


```{r}
# ==============================================================================
# 1. SETUP PARAMETERS
# ==============================================================================
n_iterations <- 10  # How many times to repeat the random split
significance <- 0.05
results_list <- list()

set.seed(123) # Optional: set seed for reproducibility of the whole sequence

# ==============================================================================
# 2. ITERATION LOOP
# ==============================================================================
for (i in 1:n_iterations) {
  
  message(paste("Running Iteration:", i, "/", n_iterations))
  
  split_info <- group_info %>%
    group_by(Genotype) %>%
    mutate(Split_Label = sample(c(1, 2), size = n(), replace = FALSE)) %>%
    ungroup()
  
  rep1_info <- split_info %>% filter(Split_Label == 1)
  rep2_info <- split_info %>% filter(Split_Label == 2)
  
  # --------------------------------------------------------------------------
  # 2. Analyze & Evaluate REPLICATE 1
  # --------------------------------------------------------------------------
  # A. Run Analysis
  rep_sel <- rep1_info 
  sample_srep <- rep_sel$ID
  raw_use <- Big_Counts[,sample_srep]
  colnames(raw_use) <- rep_sel$Genotype
  norm_use <- Big_Counts_Norm[,sample_srep]
  colnames(norm_use) <- rep_sel$Genotype
  norm_factors_use <- Big_C[sample_srep]
  names(norm_factors_use) <- rep_sel$Genotype
  res1 <- twoStageLRT_app(hybrid_info = hybrid_info, raw_use = raw_use,
                          norm_use = norm_use, norm_factors_use = norm_factors_use,
                           gene_groups_no = 3, n_fam_thres = 2, alpha0 = 0.01)
  
  # B. Evaluate (Compare against Ground Truth)
  eval1 <- evaluate_performance(pred_p  = res1$pval_NB, 
                                pred_q  = res1$qval_NB, 
                                truth_p = p_mat_rep, # Ground Truth P
                                truth_q = q_mat_rep, # Ground Truth Q
                                label   = paste0("Iter", i, "Rep1"),
                                significance = 0.05)
  eval1$Iteration <- i
  
  # --------------------------------------------------------------------------
  # 3. Analyze & Evaluate REPLICATE 2
  # --------------------------------------------------------------------------
  # A. Run Analysis
  rep_sel <- rep2_info 
  sample_srep <- rep_sel$ID
  raw_use <- Big_Counts[,sample_srep]
  colnames(raw_use) <- rep_sel$Genotype
  norm_use <- Big_Counts_Norm[,sample_srep]
  colnames(norm_use) <- rep_sel$Genotype
  norm_factors_use <- Big_C[sample_srep]
  names(norm_factors_use) <- rep_sel$Genotype
  res2 <- twoStageLRT_app(hybrid_info = hybrid_info, raw_use = raw_use,
                          norm_use = norm_use, norm_factors_use = norm_factors_use,
                           gene_groups_no = 3, n_fam_thres = 2, alpha0 = 0.01)
  
  # B. Evaluate
  eval2 <- evaluate_performance(pred_p  = res2$pval_NB, 
                                pred_q  = res2$qval_NB, 
                                truth_p = p_mat_rep, 
                                truth_q = q_mat_rep, 
                                label   = paste0("Iter", i, "Rep2"),
                                significance = 0.05)
  eval2$Iteration <- i
  
  # --------------------------------------------------------------------------
  # 4. Store Results
  # --------------------------------------------------------------------------
  results_list[[i]] <- rbind(eval1, eval2)
}

# Combine all results
final_results_df <- do.call(rbind, results_list)

# View
print(head(final_results_df))

# Summary Statistics
summary_stats <- final_results_df %>%
  group_by(Replicate_ID) %>%
  summarise(
    Mean_AUC = mean(AUC, na.rm=T),
    Mean_CCC = mean(CCC_LogP, na.rm=T),
    Mean_Precision = mean(Precision, na.rm=T)
  )

print(summary_stats)
```

