---
title: "TwoStageLRT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TwoStageLRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(TwoStageLRT)
```


This file shows the simulation in the original paper under a simple setting.

## Simulation steps

1. Simulate rate for Gamma distribution: We begin by simulating gene-wise rate parameters from a uniform distribution with range from 0.05 to 0.2 to ensure that the mean expression level for each gene remains consistent across families, but varies across different genes.

2. Simulate parental mean expression: For each gene-family combination, we simulate the mean of the female variety from a gamma distribution with a shape parameter of 30 and the simulated gene-wise rate parameter. The male mean matrix is set to be identical to the female mean matrix. 

3. Simulate hybrid mean expression: We specify which gene-family combinations exhibit MPH, i.e., the MPH pattern, and the MPH strength, i.e., MPH ratio which is defined  as $\delta_{gh}=\mu_{gh3}/(\frac{\mu_{gh1}+\mu_{gh2}}{2})-1$. Inspired by the real data example in the original paper where there is considerable variation in the proportions of MPH hybrids between different genes, we aim to simulate some genes that tend to exhibit MPH in most families and others exhibit MPH in only a few families. The percentage of MPH families is set to be either 90\% for genes tending to exhibit MPH or 5\% for genes tending  not to exhibit MPH. 
   
   Regarding the MPH ratio, we introduce positive heterosis ($\delta_{gh}>0$), negative heterosis ($\delta_{gh}<0$), and non-heterosis ($\delta_{gh}=0$) gene-family combinations into our simulation. We consider signal strength pairs: $\delta_{gh} = 2$ for positive heterosis and $\delta_{gh} = -2/3$ for negative heterosis. 
   
4. Simulate dispersion parameters: We calculate the median mid-parent mean for each gene, denoted as $mp$. The dispersion parameter is then set to be $mp^{-1/6}-0.3$ with the addition of a random sample $\epsilon$ from a normal distribution with mean 0 and standard deviation 0.002. Although rare, negative values of dispersion are replaced by 0.

5. imulate values for normalization factors: The normalization factors are simulated from a uniform distribution with range from 0.8 to 1.3. 

6. Simulate count table for female, male and hybird: Using the female/male/hybrid mean matrix, dispersion for each gene and normalization factors for each sample, we simulate the female/male/hybird count table with dimensions $G = 100 \times H = 150$ from a negative binomial distribution.

## Testing

Just for illustration purpose, 3 datasets are generated following simulation steps above and 2sLRT is applied to each of them. To compare with, in addition to 2sLRT, we also performed the naive estimation method (denoted as Estimation) which rank the gene-family combination by the absolute values of MPH ratio estimation, and Poisson likelihood ratio test (denote as Poisson) for detection of MPH. 


To conduct 2sLRT, we specify the hyper-parameters to be: thee minimum number of null families for gene dispersion estimation $N_0$ is set to be 10, and the number of gene clusters is 2.




```{r sim settings}
mu_female_dense <- "matrix(rgamma(case$n_genes*case$n_families, shape = 30, rate=runif(case$n_genes, 0.05, 0.2)),nrow=case$n_genes, ncol=case$n_families)"

effect_factor_dense <- "rbind(matrix(ifelse(rbinom(case$n_genes/4*case$n_families, 1, 0.05)==1, case$sig_pos, 0),nrow = case$n_genes/4, ncol = case$n_families), matrix(ifelse(rbinom(case$n_genes/4*case$n_families, 1, 0.05)==1, case$sig_neg, 0),nrow = case$n_genes/4, ncol = case$n_families), matrix(ifelse(rbinom(case$n_genes/4*case$n_families, 1, 0.9)==1, case$sig_pos, 0),nrow = case$n_genes/4, ncol = case$n_families), matrix(ifelse(rbinom(case$n_genes/4*case$n_families, 1, 0.9)==1, case$sig_neg, 0),nrow = case$n_genes/4, ncol = case$n_families))"

case <- list(n_genes = 100, n_families = 150,
             sig_pos = 2, sig_neg = -2/3,
             mu_female = mu_female_dense,
             mu_male = "case$mu_female", 
             effect_factor = effect_factor_dense,
             female_norm_factor = c(0.8, 1.3), 
             male_norm_factor = c(0.8, 1.3), 
             hybrid_norm_factor = c(0.8, 1.3),
             MP_DISP_func = "inverse_func", true_disp = FALSE, true_norm = FALSE,
             n_fam_thres = 10, gene_groups_no = 2, repetition = 3, plot = TRUE, filename = NULL)
print(case)

```




## Visualization

To visualize the simulation results, for 2sLRT and Poisson LRT with p-values, we present the qqplot to compare the quantiles of p-values for non-MPH gene-family combinations among all repetitions versus quantiles of Uniform(0,1) distribution. We also report the FDR averaged over repetitions changing over nominal FDR within $(0,0.2)$ and at 0.05 nominal FDR. For each of the three methods, an average Receiver Operating Characteristic (ROC) curve over repetitions is presented, along with the average partial Area Under Curve (AUC) for false positive rate within range $(0,0.2)$. In addition, to evaluate  the dispersion estimation from 2sLRT, we also present the scatter plot of estimated gene-wise dispersion parameters (averaged over datasets) versus the corresponding true values. 


```{r sim results}
if(is.character(case$mu_female)){case$mu_female <- eval(parse(text = case$mu_female))}
if(is.character(case$mu_male)){case$mu_male <- eval(parse(text = case$mu_male))}
if(is.character(case$effect_factor)){case$effect_factor <- eval(parse(text = case$effect_factor))}
res <- Hetero_pip_2sLRT(n_genes = case$n_genes, n_families = case$n_families,
                        mu_female = case$mu_female, mu_male = case$mu_male,
                        effect_factor = case$effect_factor,
                        female_norm_factor = case$female_norm_factor,
                        male_norm_factor = case$male_norm_factor,
                        hybrid_norm_factor = case$hybrid_norm_factor,
                        true_norm = case$true_norm,
                        MP_DISP_func = case$MP_DISP_func, true_disp = case$true_disp,
                        n_fam_thres = case$n_fam_thres,
                        gene_groups_no = case$gene_groups_no,
                        repetition = case$repetition, plot = case$plot,
                        filename = case$filename)
print(res)
```

